### 2020-07-03 manuscript/manuscript.rmd
Thrid round or pass to main text and Extended Fig3

Yuanning Li^1,3^, Xingxing Shen^3^, Benjamin Evans^2^, Antonis Rokas^3^, and Casey W. Dunn^1^*

^1^Department of Ecology and Evolutionary Biology, Yale University

^2^Yale Center for Research Computing, Yale University

^3^Department of Biological Science, Vanderbilt University

@@

@@

There has been considerable recent debate about the placement of the root in the animal tree of life. This question has major implications for our understanding of the earliest events in animal evolution, including the origin of the nervous system. Some phylogenetic analyses support a root that places the first split in the phylogeny of living animals between sponges and all other animals (the Porifera-sister hypothesis), and others find support for a split between comb jellies and all other animals (Ctenophora-sister). These analyses differed in a many respects, including genes considered, species considered, molecular evolution models, and software. Here we systematically explore this question under consistent conditions by synthesizing data and results from 13 previous phylogenomic analyses and performing new standardized analyses. Published studies find support for Porifera-sister in some analyses with reduced outgroup (non-animal) sampling and complex site heterogeneous models, and Ctenophora-sister under other conditions. Our new analyses are consistent with this. Though this had been interpreted by some to suggest Ctenophora-sister is an artifact of amino-acid composition bias, our analyses instead indicate that many site heterogeneous models do support Ctenophora-sister, and Porifera-sister is a result of unconstrained overfit models with a very high number of site categories. Our analyses help reconcile variation across previous studies, and the datasets and analyses consolidated here will be a useful test-platform for phylogenetic methods development.

@@

Historically, there was little debate about the root of the animal tree of life. Porifera-sister (Fig. 1) was widely accepted though rarely tested. By contrast, there has long been uncertainty about the relationship of Ctenophora to other animals [@Wallberg:2004ws]. The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores rather than sponges are our most distant living animal relative. Since then, many more relevant phylogenomic studies have been published, with some analyses that find support for Ctenophora-sister, some for Porifera-sister, and some neither (Extended Data Table 1). The extensive technical variation across these studies has been important to advancing the understanding of the sensitivity of these analyses, revealing for example that outgroup and model selection can have a large impact on result. But the extensive technical variation has also made it difficult to synthesize these results to understand the details of this sensitivity. This methodological variance is often confounded, for example when a particular model is only available in one software package.

To advance this problem it is therefore critical to synthesize published studies and explore variation in methods and results in a standardized and systematic framework. Here, we synthesize data and results from previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister from 13 studies (Extended Data Table 1). These include all such phylogenomic studies of amino acid sequence data published before 2020 for which we could obtain data matrices with gene partition annotations.

![](figures/Figure_overview.png)
**Fig. 1.** (A) The Ctenophora-sister hypothesis posits that there is a clade (designated by the orange node) that includes all animals except Ctenophora, and that Ctenophora is sister to this clade. (B) The Porifera-sister hypothesis posits that there is a clade (designated by the green node) that includes all animals except Porifera, and that Porifera is sister to this clade. Testing these hypotheses requires evaluating the support for each of these alternative nodes. (C) The animals and their outgroup choice, showing the three progressively more inclusive clades Choanimalia, Holozoa, and Opisthokonta. (D) A ctenophore. (E) A sponge.

@@

The models in the studies considered here differ according to two primary components: the exchangeability matrix $E$ and amino acid equilibrium frequencies $\Pi$. The exchangeability matrix $E$ describes the relative rates at which one amino acid changes to another. The studies considered here use exchangabilities that are the same between all amino acids (Poisson, also referred to as F81), fixed according to standard empirically estimated rates (WAG or LG), or all independently estimated from the data (GTR). The analyses considered here have site homogenous exchangeability models, with the same matrix used for all sites in all genes. The equilibrium frequencies describe the expected frequency of each amino acid, which captures the fact that some amino acids are much more common than others. The analyses differ in whether they take a homogeneous approach and jointly estimate the same frequency across all sites, or add parameters that allow heterogeneous equilibrium frequencies that differ across sites. Heterogeneous approaches include CAT [@Lartillot:2004dq], which is implemented in the software PhyloBayes and has been widely applied to questions of deep animal phylogeny. The models that are applied in practice are heavily influenced by computational costs, model availability in software, and convention. For example, on the questions considered here, CAT site-heterogeneous models of equilibrium frequency have only been used in combination with Poisson and GTR exchangeability matrices, but not LG and WAG exchangeability matrices. Model effect is further confused by the abbreviations that are used for models. Papers often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to heterogeneous equilibrium frequencies across sites and WAG to a particular exchangeability matrix. In this literature, CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation. To avoid confusion on this point, here we always specify the exchangeability matrix first (e.g., GTR), followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (e.g., CAT). If site homogeneous equilibrium frequencies are used, we refer to the exchangeability matrix alone. Gamma-rate heterogeneity, a scalar that accommodate total rate of change across sites, is used in almost every analysis conducted here and we generally omit its designation.

High-throughput sequencing allows investigators to readily assemble matrices with hundreds or thousands of protein-coding genes from a broad diversity of animal species (Extended Data Table 1). Studies of animal phylogeny have used a wide variety of different approaches to identifying and selecting genes for their matrices. As a result, the paritcular genes selected for analysis vary widely (Fig. 2, Extended Data Fig. 1). The composition of of these genes varies in several ways, including the fraction of BUSCO and ribosomal protein genes in the matrix (Extended Data Fig. 2).

@@

**Fig. 2.** An overview of previous studies. (A) A time-series showing the main conclusions of previous phylogenomic studies that indicated their analyses supported Ctenophora-sister ( orange nodes) or Porifera-sister (green nodes). (B) Each of the primary matrices considered here, color coded by taxon sampling. Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.

Ingroup taxon sampling also varies widely between studies (Fig. 2, Extended Data Fig. 1). Sampling of ingroup taxa (animals) was initially biased toward Bilateria. Sampling of non-bilaterian animals, including sponges and ctenophores, has improved over time as studies add additional taxa from these poorly studied groups (Extended Data Fig. 1). Within each clade, there is often considerable variation in taxon sampling and therefore often little species overlap across studies (Extended Data Fig. 1). This variation is in part because newer sequencing technologies in more recent papers are usually not applied to the exact same species that were in earlier papers.

Sampling of outgroup taxa (non-animals, in this case) is critical to phylogenetic rooting questions, since the node where the outgroup subtree attaches to the ingroup subtree is the root of the ingroup. There has therefore been extensive focus on improving outgroup sampling when testing phylogenetic hypotheses about rooting [@graham2002rooting]. Most studies addressing the animal root have removed more distantly related outgroup taxa in some analyses to explore the effect of outgroup selection to ingroup topology [ @ryan2013genome; @pisani2015genomic]. Three progressively more inclusive clades have often been investigated: Choanimalia (animals plus most closely related Choanoflagellatea), Holozoa (Choanimalia plus more distantly related Holozoa) and Opisthokonta (Holozoa + Fungi).

@@

**Fig. 3.**  A summary of phylogenomic results from the previous studies and reanalysis conducted in this study. Each point represents a phylogenetic Analyses with a support of bootstrap values less than `r bootstrap_threshold`% or a posterior probability less than `r posterior_prob_threshold`% are considered as unresolved. The clades are orgainized with increased outgroup sampling higher on the vertical axis, and the models are organized in general by increasing complexity to the right on the horizontal axis. (A) Analyses transcribed from the literature (related to Supplementary Table 2). (B) New phylogenomic analyses conducted in this study (related to Supplementary Table 3).

We parsed `r analyses_processed %>% filter(status=="published") %>% nrow()` previous phylogenetic analyses from `r analyses_processed %>% filter(status=="published") %>% pull (manuscript) %>% unique() %>% length()` studies (Fig. 3A and Supplementary Table 2). The conclusions of five studies strongly favor of Porifera-sister and 10 favor Ctenophora-sister (Extended Data Table 1). Three studies are based entirely on previously published data, and the remainder add data for one or more species.

@@

One of the challenges of interpreting support for the placement of the animal root across published studies is that different programs, software versions, and settings have been used across analyses. This extensive variation can make it difficult to identify the primary factors that lead to different results. Here we first reanalyze the primary matrices from each study under consistent conditions with IQ-TREE [@nguyen2014iq] with multiple evolutionary models. We selected this tool because it has greater model flexibility than most other phylogenetic tools [@zhou2017evaluating] (Fig. 3B; Supplementary Table 3). Importantly, it has C10-C60 equilibrium frequencies [@si2008empirical] that, like CAT, allow models to accommodate heterogeneity across sites.

@@

For each of the published studies, we selected the matrix that was the primary focus of the manuscript for further analysis. For each of these matrices, we progressively trimmed taxon sampling to create Opisthokonta, Holozoa, and Choanimalia versions, where permitted by original outgroup sampling. This produced `r analyses_processed %>% filter(status=="new") %>% nrow()` matrices.

For all but the three largest matrices, we tested the relative fit of a variety of models, both with and without C60 accommodation of site heterogeneity in equilibrium frequencies, using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQ-TREE. In all cases, models with C60 fit these matrices better than the site-homogeneous models. We then inferred support under the best-fit model (Extended Data Table 2), except for the three largest matrices where we used LG+C60. We then analyzed each matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG, GTR and Poisson+C60 (Extended Data Table 2).

@@

Our new analyses show that, with restricted outgroup sampling, analyses of the same matrices with two deifferent means of accommodating site heterogeneity in equilibrium frequencies (C60 in IQ-TREE and CAT in PhyloBayes) get different results. This indicates that the traditional framing of the problem, that accommodating rate heterogeneity leads to support for Porifera-sister, is not correct. Instead, there is something about PhyloBayes CAT analyses specifically that leads to support for Porifera-sister.

@@

cat_categories$n_categories_poisson = cat_categories$`Infered categories from Poisson-CAT`
cat_categories$n_categories_gtr = cat_categories$`Infered categories from GTR-CAT`

@@

A primary difference between the C (like C60) and CAT site-heterogeneous models is the number of equilibrium frequency categories. CAT models employ a Dirichlet process prior to infer the number of equilibrium frequency categories, so the number of categories is variable [@Lartillot:2004dq]. IQ-TREE implements C models [@si2008empirical] with a fixed number of categories, from C10 to C60. Differences in analysis results could therefore be due to differences in the number of categories. The number of categories inferred by CAT in PhyloBayes can be very high (Supplementary Table 4), with a mean here of `r cat_categories$n_categories_poisson %>% mean() %>% round(1)` categories for Poisson+CAT analyses and `r cat_categories$n_categories_gtr %>% mean() %>% round(1)` categories for GTR+CAT analyses. This requires a very large number of additional parameters that must be estimated from the data. When ModelFinder is used to select the number of categories in IQ-TREE C models, it often selects a number far less than 60, suggesting that 60 is already too high according to ModelFinder (Supplementary Table 5). These findings are consistent with earlier concern that CAT models can overestimate the number of equilibrium frequency categories in simulated matrices [@whelan2016let].

@@

Published studies have found that the GTR-CAT model often has a better fit than Poisson-CAT, and investigators have suggested that it is more appropriate than the Poisson-CAT model for this question (XX citations). Due to computational limitations of GTR models, we only ran GTR+CAT and GTR+nCAT60 models on matrices with Choanimalia sampling. We found that for both Whelan2017 and Philippe matrices shifted from Porifera-sister with Poisson-CAT model to Ctenophora-sister using GTR-CAT model. Moreover, we also found all results strongly supported Ctenophora-sister with GTR-nCAT60 models.

@@

```
**Fig. 4.**  Sensitive analyses with representative data matrices analyzed by different number of equilibrium frequency categories (nCAT) in PhyloBayes. Statistical support values (posterior probabilities) were obtained from three data matricies using the site-heterogeneous Poisson-CAT model with different categories” (A) Phlippe2009_Choanozoa; (B) Philippe2009_Holozoa; (C) Whelan2017_full_Choanozoa; (D) Whelan2017_strict. Statistical support for Ctenophora-sister and Porifera-sister is indicated in orange and green, respectively. Support values are from the sensitive analyses are shown in Supplementary Table 6.

@@

To further explore the variation described above, we quantify the phylogenetic signal for Porifera-sister and Ctenophora-sister topologies across three representative data matrices when varying outgroup sampling and model (Extended Data Fig. 3). By calculating differences in log-likelihood scores for these topologies for every gene ($\Delta$|lnL|) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models. Moreover, the outgroup choice has little impact on the distribution of the support of phylogenetic signal in analyses with site-homogeneous models. This finding is largely consistent with the the previously observed distribution of support for Ctenophora-sister in other data matrices [@shen2017contentious].

Although a higher proportion of genes support Ctenophora-sister with site-heterogeneous C60 models, the phylogenetic signal decreases in many genes using C60 models compared to site-homogeneous models. In an extreme case, in matrices from Ryan2013_est nearly 30% of genes changed from strong to week Ctenophora-sister signal ($\Delta$|lnL|<2) (Extended Data Fig. 3; Supplementary Table 7). In contrast to the C60 models, the phylogenetic signal largely increased in Poisson-CAT models, and the outgroup choice has a major effect of the distribution of phylogenetic signal (Extended Data Fig. 3). We found that the number of genes that support Ctenophora-sister in analyses with CAT decreases from XXs in matrices with distant outgroups (Opisthokonta) to XXb when outgroups are restricted (Choanimalia, Supplementary Table 7).

@@

Other approaches have been taken to addressing heterogeneity in equilibrium frequencies. For example, Feuda *et al.* [@Feuda:2017ew] recoded the full set of twenty amino acids into six groups of amino acids. These groups tend to have more frequent evolutionary changes within them than between them [@Susko:2007ds]. Recoding could, like CAT and C models, address variation across sites, but it could also accommodate variation across lineages, and it was suggested that this approach favors Porifera-sister [@Feuda:2017ew]. Our own analyses (Supplementary text, Supplementary Fig. XX-XX) indicate, however that the impact of data recoding is is largely due to discarding information, not accommodating variation in amino acid composition. These findings indicate that recoding can be a problematic method for addressing heterogeneity.

@@

Resolving the placement of the root in the animal tree of life has been a very difficult phylogenetic problem. By synthesizing past phylogenomic studies and performing new analyses, we confirm that support of Porifera-sister is only recovered by site-heterogeneous CAT models with restricted outgroup sampling, and then only in some of such analyses. This demonstrates an important interaction between model selection and taxon sampling. Through controlled analyses we are able to identify the specific aspect of the models that is involved in this variation -- the number of categories used to accommodate site heterogeneity in equilibrium frequency. Site homogenous models have a single category that is applied to all sites. C models, as implemented in IQ-TREE, have 10-16 categories. CAT models can estimate the number of categories, in which case they use more than 500 for the matrices assessed here, or set to a specific number with nCAT. When outgroup sampling is fixed to a restricted subset (a requirement for any analysis to support Porifera-sister), root position is dependent on the number of categories. As the number of categories is increased, at some point support for Ctenophora-sister diminishes and support for Porifera-sister increases (Fig. 4). As part of this process, the number of genes with phylogenetic signal for each hypothesis also changes.

@@

Overcoming the challenges that face this difficult and important question goes beyond models. In particular, there are many genomic data that are highly relevant to this problem that have not yet been collected and analyzed in a phylogenetic context. In particular, there are very few chromosome-level genome assembles outside of Bilateria. There will be important advances in genome sequencing in coming years that will have a major impact on this problem. This is is about more than more data - high quality chromosome-level assemblies will greatly facilitate gene homology assessments made during matrix construction and reduce the possibility of contamination. We hope that the work we have conducted here to consolidate many datasets and analyses in standard formats will make it easier for other investigators to engage in this particularly interesting difficult phylogenetic problem, and also to develop methods and tools that will help with other difficult phylogenetic problems.

@@

All tree files, intermidiate results and scripts/commands associated with this study are available at https://github.com/caseywdunn/animal_root, doi will be provided upon publicaton.

@@

#### Outgroup taxa sampling with C60 and CAT models
Because different choices of outgroups could also affect phylogenetic inference as suggested in previous analyses, we parsed the full data matrices into three different types of outgroups: Choanimalia , Holozoa and Opisthokonta. These datasets include the same set of genes but differ in the composition of outgroup species. Choanimalia only includes choanofagellates as outgroup; Holozoa also includes more distantly related holozoans; Opistokonta also includes Fungi. For each Choanimalia data matrice, both C60 models in IQ-TREE and Poisson-CAT models in PhyloBayes were conducted. The maximum likelihood analysis was performed using the best-fit substitution model identified as above and nodal support was assessed with 1000 ultrafast bootstrap replicates using IQ-TREE. Moreover, bayesian inference with the site-heterogeneous poisson-CAT model was done with PhyloBayes-MPI v1.8. To minimize computational burden, CAT-GTR models were only performed in the representative Chaonimalia matrices from Philippe2009, Ryan2013_est and Whelan2017_full.

>>>>

Yuanning Li^1,3^, Xing-xing Shen^2,3^, Benjamin Evans^4^, Steven H.D. Haddock^5^, Antonis Rokas^3^ and Casey W. Dunn^1^*

^1^Department of Ecology and Evolutionary Biology, Yale University, New Haven, CT 06511, USA

^2^Department of Agriculture, Zhejiang University, Hangzhou, Zhejiang, China

^3^Department of Biological Science, Vanderbilt University, Nashville 37235, USA

^4^Yale Center for Research Computing, Yale University, New Haven, CT 06511, USA

^5^Monterey Bay Aquarium Research Institute, Moss Landing, CA 95039, USA

@@

ORCIDs:

Yuanning Li: 0000-0001-5765-1419
Xing-Xing Shen: 0000-0002-8436-595X
Benjamin Evans:
Steven H.D. Haddock: 0000-0001-9420-4482
Antonis Rokas: 0000-0002-7248-6551
Casey W. Dunn: 0000-0003-0628-5150

@@

There has been considerable recent debate about the placement of the root in the animal tree of life. This question has major implications for our understanding of the earliest events in animal evolution, including the origin of the nervous system. Some phylogenetic analyses support a root that places the first split in the phylogeny of living animals between sponges and all other animals (the Porifera-sister hypothesis), and others find support for a split between comb jellies and all other animals (Ctenophora-sister). These analyses differed in many respects, including genes considered, species considered, molecular evolution models, and software. Here we systematically explore this question under consistent conditions by synthesizing data and results from 15 previous phylogenomic analyses and performing new standardized analyses. Published studies find support for Porifera-sister in some analyses with reduced outgroup (non-animal) sampling and complex site heterogeneous models, and Ctenophora-sister under other conditions. Our new analyses are consistent with this. Though this had been interpreted by some to suggest Ctenophora-sister is an artifact of amino-acid composition bias, our analyses instead indicate that many site heterogeneous models do support Ctenophora-sister, and Porifera-sister is a result of unconstrained overfit models with a very high number of site categories. Our analyses help reconcile variation across previous studies, and the datasets and analyses consolidated here will be a useful test-platform for development of phylogenomic methods.

@@

Historically, there was little debate about the root of the animal tree of life. Porifera-sister (Figs. 1B, 1D) was widely accepted though rarely tested. By contrast, there has long been uncertainty about the relationship of Ctenophora to other animals [@Wallberg:2004ws]. The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores rather than sponges are our most distant living animal relative (Figs. 1A, 1C). Since then, many more relevant phylogenomic studies have been published, with some analyses that find support for Ctenophora-sister, some for Porifera-sister, and some neither (Extended Data Table 1). The extensive technical variation across these studies has been important to advancing the understanding of the sensitivity of these analyses, revealing for example that outgroup and model selection can have a large impact on result (Fig. 1E). But the extensive technical variation has also made it difficult to synthesize these results to understand the details of this sensitivity. This methodological variance is often confounded, for example when a particular model is only available in one software package.

To advance this problem it is therefore critical to synthesize published studies and explore variation in methods and results in a standardized and systematic framework. Here, we synthesize data and results from previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister from 15 studies (Fig. 2A, Extended Data Table 1). These include all such phylogenomic studies of amino acid sequence data published before 2020 for which we could obtain data matrices with gene partition annotations.

![](figures/Figure_overview.png)
**Fig. 1.** Two alternative phylogenetic hypotheses from previous phylogenomic studies. (A) The Ctenophora-sister hypothesis posits that there is a clade (designated by the orange node) that includes all animals except Ctenophora, and that Ctenophora is sister to this clade. (B) The Porifera-sister hypothesis posits that there is a clade (designated by the green node) that includes all animals except Porifera, and that Porifera is sister to this clade. Testing these hypotheses requires evaluating the support for each of these alternative nodes. (C) The animals and their outgroup choice, showing the three progressively more inclusive clades Choanimalia, Holozoa, and Opisthokonta. (D) A ctenophore. (E) A sponge.

@@

The models in the studies considered here differ according to two primary components: the exchangeability matrix $E$ and amino acid equilibrium frequencies $\Pi$. The exchangeability matrix $E$ describes the relative rates at which one amino acid changes to another. The studies considered here use exchangabilities that are the same between all amino acids (Poisson, also referred to as F81), fixed according to standard empirically estimated rates (WAG or LG), or all independently estimated from the data (GTR). The analyses considered here have site homogenous exchangeability models (site-homogeneous model), with the same matrix used for all sites in all genes. The equilibrium frequencies describe the expected frequency of each amino acid, which captures the fact that some amino acids are much more common than others. The analyses differ in whether they take a homogeneous approach and jointly estimate the same frequency across all sites, or add parameters that allow heterogeneous equilibrium frequencies that differ across sites. Heterogeneous approaches include CAT [@Lartillot:2004dq], which is implemented in the software PhyloBayes and has been widely applied to questions of deep animal phylogeny. The models that are applied in practice are heavily influenced by computational costs, model availability in software, and convention. For example, on the questions considered here, CAT site-heterogeneous models of equilibrium frequency have only been used in combination with Poisson and GTR exchangeability matrices, but not LG and WAG exchangeability matrices. Model effect is further confused by the abbreviations that are used for models. Papers often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to heterogeneous equilibrium frequencies across sites and WAG to a particular exchangeability matrix. In this literature, CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation. To avoid confusion on this point, here we always specify the exchangeability matrix first (e.g., GTR), followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (e.g., CAT). If site homogeneous equilibrium frequencies are used, we refer to the exchangeability matrix alone. Gamma-rate heterogeneity, a scalar that accommodates total rate of change across sites, is used in almost every analysis conducted here and we generally omit its designation.

High-throughput sequencing allows investigators to readily assemble matrices with hundreds or thousands of protein-coding genes from a broad diversity of animal species (Extended Data Table 1). Studies of animal phylogeny have used a wide variety of different approaches to identifying and selecting genes for their matrices. As a result, the particular genes selected for analysis vary widely (Fig. 2B, Extended Data Fig. 1). The composition of these genes varies in several ways, including the fraction of BUSCO and ribosomal protein genes in the matrix (Extended Data Fig. 2).

@@

**Fig. 2.** An overview of previous phylogenomic studies. (A) A time-series showing the main conclusions of previous phylogenomic studies that indicated their analyses supported Ctenophora-sister (orange nodes) or Porifera-sister (green nodes) based on their main conclusion from each study. (B) Each of the primary matrices considered here, color coded by taxon sampling. Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.

Ingroup taxon sampling also varies widely between studies (Fig. 2B, Extended Data Fig. 1). Sampling of ingroup taxa (animals) was initially biased toward Bilateria. Sampling of non-bilaterian animals, including sponges and ctenophores, has improved over time as studies add additional taxa from these poorly studied groups (Extended Data Fig. 1). Within each clade, there is often considerable variation in taxon sampling and therefore often little species overlap across studies (Extended Data Fig. 1). This variation is in part because newer sequencing technologies in more recent papers are usually not applied to the exact same species that were in earlier papers.

Sampling of outgroup taxa (non-animals, in this case) is critical to phylogenetic rooting questions, since the node where the outgroup subtree attaches to the ingroup subtree is the root of the ingroup. There has therefore been extensive focus on improving outgroup sampling when testing phylogenetic hypotheses about rooting [@graham2002rooting]. Most studies addressing the animal root have removed more distantly related outgroup taxa in some analyses to explore the effect of outgroup selection to ingroup topology [@ryan2013genome; @pisani2015genomic]. Three progressively more inclusive clades have often been investigated: Choanimalia (animals plus most closely related Choanoflagellatea), Holozoa (Choanimalia plus more distantly related Holozoa) and Opisthokonta (Holozoa + Fungi).

@@

**Fig. 3.**  A summary of phylogenomic results from the previous studies and reanalysis conducted in this study. Each point represents a phylogenetic analyses with a support of bootstrap values less than `r bootstrap_threshold`% or a posterior probability less than `r posterior_prob_threshold`% are considered as unresolved. The clades are organized with increased outgroup sampling higher on the vertical axis, and the models are organized in general by increasing complexity to the right on the horizontal axis. (A) Analyses transcribed from the literature (related to Supplementary Table 2). (B) New phylogenomic analyses conducted in this study (related to Supplementary Table 3).

We parsed `r analyses_processed %>% filter(status=="published") %>% nrow()` previous phylogenetic analyses from `r analyses_processed %>% filter(status=="published") %>% pull (manuscript) %>% unique() %>% length()` studies (Fig. 3A and Supplementary Table 2). The conclusions of five studies strongly favor Porifera-sister and 10 favor Ctenophora-sister (Extended Data Table 1). Three studies are based entirely on previously published data, and the remainder add data for one or more species.

@@

One of the challenges of interpreting support for the placement of the animal root across published studies is that different programs, software versions, and settings have been used across analyses. This extensive variation can make it difficult to identify the primary factors that lead to different results. Here we first reanalyze the primary matrices from each study under consistent conditions with IQ-TREE [@nguyen2014iq] with multiple evolutionary models. We selected this tool because it has greater model flexibility than most other phylogenetic tools [@zhou2017evaluating] (Fig. 3B; Supplementary Table 3). Importantly, it has C models (C10-C60 equilibrium frequencies) [@si2008empirical] that, like CAT, allow models to accommodate heterogeneity across sites.

@@

Among 15 previous studies,  three studies [@pisani2015genomic; @whelan2016let; @feuda2017improved]are based entirely on previously published data and gene-partition data is not available from one study [@Pick:2010eb]. For each of the published studies, we selected the matrix that was the primary focus of the manuscript for further analysis. For each of these matrices, we progressively trimmed taxon sampling to create Opisthokonta, Holozoa, and Choanimalia versions, where permitted by original outgroup sampling. This produced `r analyses_new$matrix%>% unique()%>% length()`from 11 studies.

For all but the three largest matrices, we tested the relative fit of a variety of models, both with and without C10-C60 accommodation of site heterogeneity in equilibrium frequencies, using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQ-TREE. In all cases, models with C60 fit these matrices better than the site-homogeneous models. We then inferred support under the best-fit model (Extended Data Table 2), except for the three largest matrices where we used LG+C60. We then analyzed each matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG, GTR and Poisson+C60 (Extended Data Table 2).

@@

Our new analyses show that, with restricted outgroup sampling, analyses of the same matrices with two different means of accommodating site heterogeneity in equilibrium frequencies (C60 in IQ-TREE and CAT in PhyloBayes) get different results. This indicates that the traditional framing of the problem, that accommodating rate heterogeneity leads to support for Porifera-sister, is not correct. Instead, there is something about PhyloBayes CAT analyses specifically that leads to support for Porifera-sister.

@@

cat_categories$n_categories_poisson = cat_categories$`Infered categories from Poisson+CAT`
cat_categories$n_categories_gtr = cat_categories$`Infered categories from GTR+CAT`

@@

A primary difference between the C (like C60) and CAT site-heterogeneous models is the number of equilibrium frequency categories. CAT models employ a Dirichlet process prior to infer the number of equilibrium frequency categories, so the number of categories is variable [@Lartillot:2004dq]. IQ-TREE implements C models [@si2008empirical] with a fixed number of categories, from C10 to C60. Differences in analysis results could therefore be due to differences in the number of categories. The number of categories inferred by CAT in PhyloBayes can be very high (Supplementary Table 4), with a mean here of `r cat_categories$n_categories_poisson %>% mean() %>% round(1)` categories for Poisson+CAT analyses and `r cat_categories$n_categories_gtr %>% mean() %>% round(1)` categories for GTR+CAT analyses. This requires a very large number of additional parameters that must be estimated from the data. Despite C60 models have a better fit than site-homogeneous mdoels in ModelFinder, the program reported that the C60 model might be overfitting because some category frequencies are estimated close to zero (except the Whelan2015, Whelan2017 and Simion2017 matrices), suggesting that 60 is probabaly already too high according to ModelFinder (Supplementary Table 5). These findings are consistent with earlier concern that CAT models can overestimate the number of equilibrium frequency categories in simulated matrices [@whelan2016let].

@@

Published studies have found that the GTR+CAT model often has a better fit than Poisson+CAT, and investigators have suggested that it is more appropriate than the Poisson+CAT model for this question [@simion2017large]. Due to computational limitations of GTR models, we only ran GTR+CAT and GTR+nCAT60 models on representative matrices with Choanimalia sampling. We found that for both Whelan2017 and Philippe matrices shifted from Porifera-sister with Poisson+CAT model to Ctenophora-sister using GTR+CAT model. Moreover, we also found all results strongly supported Ctenophora-sister with GTR+nCAT60 models.

@@

```
**Fig. 4.**  Sensitive analyses with representative data matrices analyzed by different number of equilibrium frequency categories (nCAT) in PhyloBayes. Statistical support values (posterior probabilities) were obtained from three data matrices using the site-heterogeneous Poisson+CAT model with different categories. (A) Phlippe2009_Choanozoa; (B) Philippe2009_Holozoa; (C) Whelan2017_full_Choanozoa; (D) Whelan2017_strict. Statistical support for Ctenophora-sister and Porifera-sister is indicated in orange and green, respectively. Support values are from the sensitive analyses are shown in Supplementary Table 6.

@@

To further explore the variation described above, we quantify the phylogenetic signal for Porifera-sister and Ctenophora-sister topologies across three representative data matrices when varying outgroup sampling and model (Extended Data Fig. 3). By calculating differences in log-likelihood scores for these topologies for every gene ($\Delta$|lnL|) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models. Moreover, the outgroup choice has little impact on the distribution of the support of phylogenetic signal in analyses with site-homogeneous models. This finding is largely consistent with the previously observed distribution of support for Ctenophora-sister in other data matrices [@shen2017contentious].

Although a higher proportion of genes support Ctenophora-sister with site-heterogeneous C60 models, the phylogenetic signal decreases in many genes using C60 models compared to site-homogeneous models. In an extreme case, in matrices from Ryan2013_est nearly 30% of genes changed from strong to week Ctenophora-sister signal ($\Delta$|lnL|<2) (Extended Data Fig. 3; Supplementary Table 7). In contrast to the C60 models, the phylogenetic signal largely increased in Poisson+CAT models, and the outgroup choice has a major effect of the distribution of phylogenetic signal (Extended Data Fig. 3). For example in Whelan2017_full matrix, we found that the number of genes that support Ctenophora-sister in analyses with CAT decreases from 57.5% in matrices with distant outgroups (Holozoa) to 35.4 when outgroups are restricted (Choanimalia, Supplementary Table 7).

@@

Other approaches have been taken to addressing heterogeneity in equilibrium frequencies. For example, Feuda *et al.* [@Feuda:2017ew] recoded the full set of twenty amino acids into six groups of amino acids. These groups tend to have more frequent evolutionary changes within them than between them [@Susko:2007ds]. Recoding could, like CAT and C models, address variation across sites, but it could also accommodate variation across lineages, and it was suggested that this approach favors Porifera-sister [@Feuda:2017ew]. Our own analyses (Supplementary text, Supplementary Figs. 4-6) indicate, however that the impact of data recoding is is largely due to discarding information, not accommodating variation in amino acid composition. These findings indicate that recoding can be a problematic method for addressing heterogeneity.

@@

Resolving the placement of the root in the animal tree of life has been a very difficult phylogenetic problem. By synthesizing past phylogenomic studies and performing new analyses, we confirm that support of Porifera-sister is only recovered by site-heterogeneous CAT models with restricted outgroup sampling, and then only in some of such analyses. This demonstrates an important interaction between model selection and taxon sampling. Through controlled analyses we are able to identify the specific aspect of the models that is involved in this variation -- the number of categories used to accommodate site heterogeneity in equilibrium frequency. Site-homogeneous models have a single category that is applied to all sites. C models, as implemented in IQ-TREE, have 10-60 categories. CAT models can estimate the number of categories, in which case they use more than 500 for the matrices assessed here, or set to a specific number with nCAT. When outgroup sampling is fixed to a restricted subset (a requirement for any analysis to support Porifera-sister), root position is dependent on the number of categories. As the number of categories is increased, at some point support for Ctenophora-sister diminishes and support for Porifera-sister increases (Fig. 4). As part of this process, the number of genes with phylogenetic signal for each hypothesis also changes.

@@

Overcoming the challenges that face this difficult and important question goes beyond models. In particular, there are many genomic data that are highly relevant to this problem that have not yet been collected and analyzed in a phylogenetic context. In particular, there are very few chromosome-level genome assemblies outside of Bilateria. There will be important advances in genome sequencing in coming years that will have a major impact on this problem. This is about more than more data - high quality chromosome-level assemblies will greatly facilitate gene homology assessments made during matrix construction and reduce the possibility of contamination. We hope that the work we have conducted here to consolidate many datasets and analyses in standard formats will make it easier for other investigators to engage in this particularly interesting difficult phylogenetic problem, and also to develop methods and tools that will help with other difficult phylogenetic problems.

@@

All tree files, intermediate results and scripts/commands associated with this study are available at https://github.com/caseywdunn/animal_root, doi will be provided upon publication.

@@

#### Outgroup taxa sampling with C60 and CAT models
Because different choices of outgroups could also affect phylogenetic inference as suggested in previous analyses, we parsed the full data matrices into three different types of outgroups: Choanimalia , Holozoa and Opisthokonta. These datasets include the same set of genes but differ in the composition of outgroup species. Choanimalia only includes choanofagellates as outgroup; Holozoa also includes more distantly related holozoans; Opistokonta also includes Fungi. For each Choanimalia data matrice, both C60 models in IQ-TREE and Poisson+CAT models in PhyloBayes were conducted. The maximum likelihood analysis was performed using the best-fit substitution model identified as above and nodal support was assessed with 1000 ultrafast bootstrap replicates using IQ-TREE. Moreover, bayesian inference with the site-heterogeneous Poisson+CAT model was done with PhyloBayes-MPI v1.8. To minimize computational burden, GTR+CAT models were only performed in the representative Chaonimalia matrices from Philippe2009, Ryan2013_est and Whelan2017_full.
